Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-04 Paper/method.tex
..

In this section, we present the computational strategy used for computing the active subspace in the case of
H$_2$/O$_2$ reaction kinetics with uncertain pre-exponent of individual reaction rates provided in
Table~\ref{tab:kinetics}. Specifically, we employ two strategies: One that involves computation of model
gradients in order to construct the matrix (grad-based), $\hat{\mat{C}}$ in~\eqref{eq:chat}, and the other that involves a local 
linear approximation of the model output to circumvent the computational cost associated with gradient
computation (grad-free). The grad-based approach is based on Algorithm 1.1, and the grad-free approach is
based on Algorithm 1.2 in~\cite{Constantine:2015}. However, in this work, we have implemented the
two strategies in an iterative manner and have demonstrated potential for additional savings in both cases.
Additionally, for the present application, our findings based on the two strategies are observed to be consistent.
Since, the grad-free approach is relatively more efficient, this observation encourages its use for high-dimensional
applications as discussed later in Section~\ref{sec:app}. Note however, that the model output is required to be
differentiable throughout the input domain in both cases.

\subsection{Grad-based approach} (\hayleynote{Better to use gradient rather than grad})
\label{sub:grad}

As discussed earlier, in the grad-based approach, the elements of $\hat{\mat{C}}$ are estimated using model gradients.
 In situations where the gradient is not available analytically, one could consider using
finite difference and other relatively more efficient techniques such as automatic differentiation~\cite{Kiparissides:2009} 
and adjoint-based gradient computation~\cite{Jameson:1988,Gunzburger:2003,Borzi:2011,Alexanderian:2017}.
Model evaluations at neighbouring points are required if finite difference is used. Hence, for 
$N$ samples in a $d$-dimensional parameter space, $N(d+1)$ model evaluations are needed.

We begin by evaluating the gradient of the model output, $\nabla_{\bm{\xi}}f$, at an initial set of $n_1$ samples. 
For this purpose, $\bm{\xi}_i$'s are projected to the physical space as $\bm{\theta}_i$'s. Using the gradient
evaluations, the matrix, $\hat{\mat{C}}$ is computed. Eigenvalue decomposition of $\hat{\mat{C}}$ yields an initial
estimate of the dominant eigenspace, $\hat{\bm{w}_1}$ and corresponding set of eigenvalues, $\hat{\bm{\Lambda}_1}$.
At each subsequent iteration, model evaluations are generated at a new set of $n_s$ samples. The new set
of gradient evaluations are augmented with the available set to re-construct $\hat{\mat{C}}$ followed by its eigenvalue
decomposition. The relative
change in the norm of the difference in squared value of individual components of the dominant eigenvectors between subsequent 
iterations is evaluated. The process is terminated and the resulting eigenspace is considered to be converged once
highest relative change at iteration $r$, $\delta w_{1,j}^{(r)}$, is smaller a given tolerance, $\tau$. A regression fit to
 $G(\bm{w}_1^\top\bm{\xi})$ is used as a surrogate to characterize and quantify the uncertainty in the model
 output. Moreover, the components of the dominant eigenvectors are used to compute the activity scores which
 provide an insight into the relative importance of the uncertain parameters. The sequence of steps for the grad-based
 approach are provided in Algorithm~\ref{alg:grad}.

@@

\bigskip
\begin{breakablealgorithm}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
  \caption{An iterative gradient-based approach for discovering the active subspace}
  \begin{algorithmic}[1]
\Require $\theta_l$, $\theta_u$, $\tau$. 
\Ensure $\Lambda$, $W$, $\alpha$ %$\eta$. %
    \Procedure{Gradient-based}{}
    \State Set $k$ = 0
	\State Draw $n_k$ random samples, $\{\bm{\xi}_i\}_{i=1}^{n_k}$ 
         according to $\pi$. 
    \State Set $N_\text{total}$ = $n_k$ 
	\iffalse \State Project to the physical space:
        $\{\bm{\theta}_k\}_{k=1}^{n_r}=\theta_l+0.5(\theta_u-\theta_l)\{\bm{\xi}_k\}_{k=1}^{n_r}$ \fi
	\State For each $i=1, \ldots, N_\text{total}$, compute $f(\bm{\xi}_i)$ and the gradient $\bm{g}^i = \nabla_{\bm{\xi}}f(\bm{\xi}_i)$
    \iffalse  
	\Statex\hspace{5mm} Using Finite Difference: 
	\Statex\hspace{5mm} i. Assign a small increment, $d\xi$.
	\Statex\hspace{5mm} ii. Augment the set of samples with neighboring points.
	\be \{\bm{\Xi}_k\}_{k=1}^{n_1(N_p+1)}:~\{\bm{\xi}_k\}_{k=1}^{n_1} \cup
        \{\xi_{k,j}+d\xi\}_{j=1}^{N_p} \nonumber
	\ee
	\Statex\hspace{5mm} iii. Project to the physical space:
        $\{\bm{\theta}_k\}_{k=1}^{n_1(N_p+1)}=\theta_l+0.5(\theta_u-\theta_l)\{\bm{\Xi}_k\}_{k=1}^{n_1(N_p+1)}$
	\Statex\hspace{5mm} iv. Using the augmented set, $\{\bm{\theta}_k\}_{k=1}^{n_1(N_p+1)}$
         compute $\bm{g}^i$.
         \fi 
	\State Compute $\mat{C}$ and its eigenvalue decomposition 
		$\mat{C}$= $\frac{1}{N_\text{total}}\sum\limits_{i=1}^{N_\text{total}}[\bm{g}^i][\bm{g}^i]^\top = W^{(k)}\Lambda^{(k)} W^{(k)\top}$
	%\State Eigenvalue decomposition, $\mat{C}$ = $W^{(0)}\Lambda^{(0)} W^{(0)\top}$%
	\State Partition: $\Lambda^{(k)}=
        \begin{bmatrix} \Lambda_1^{(k)} & \\ & \Lambda_2^{(k)} \end{bmatrix}$, 
        $W^{(k)}=\begin{bmatrix} W_1^{(k)} & W_2^{(k)} \end{bmatrix}$, 
        $\Lambda_1^{(k)}\in \mathbb{R}^{N_p\times\mathcal{S}}$
	\Loop
		\State Set $k$ = $k$ + 1
		\State Draw $n_k =  \beta n_{k-1}$  new random samples 
                $\{\bm{\xi}_i\}_{i=1}^{n_k}$  $\beta =??$  $k = n_{r-1}+1,\ldots,n_{r-1}+n_r$.

	%	\State Project $\bm{\xi}_k$~$\rightarrow$~$\bm{\theta}_k$.%
		\State Set $N_\text{total}$ = $N_\text{total}$ + $n_k$ 
		\State Compute $\bm{g}^k = \nabla_{\bm{\xi}_i}f(\bm{\xi}_i)$, 
             	$k=n_{r-1}+1, \ldots, n_{r-1}+n_r$.  
		\State Compute $\mat{C}$ = 
        	$\frac{1}{N_\text{total}}\sum\limits_{k=1}^{N_\text{total}}[\bm{g}^k][\bm{g}^k]^\top$
		\State Eigenvalue decomposition, $\mat{C}$ = $W^{(r)}\Lambda^{(r)} W^{(r)\top}$
		\State Repeat steps 5-8 above
		\State Compute $\delta W_{1,j}^{(r)}$ = 
                       \scalebox{1.25}{$\frac{\|(W_{1,j}^{r})^2 - 
                       (W_{1,j}^{r-1})^2\|}{\|(W_{1,j}^{r-1})^2\|}$}, 
                       $j = 1,\ldots,\mathcal{S}$.
		\If {$\max\left(\delta W_{1,j}^{(r)}\right)<\tau$}
			\State break
		\EndIf
	\EndLoop
	\State Compute $\alpha_i(N_\text{total}) = \sum\limits_{j=1}^{\mathcal{S}} \Lambda_{1_{i,j}}W_{1_{i,j}}^2$,
	$i=1,\ldots,d$.

@@

\subsection{Grad-free approach}
\label{sub:gradfree}

As mentioned earlier, the grad-free approach does not involve computation of the gradient of the model
output. Instead, a least-squares fit is performed to the local linear regression-based approximation
to the model output. While the grad-free approach offers a significant potential for additional 
computational savings, its ability to reasonably approximate the active subspace 
has not been investigated for a wide variety of problems~\cite{Constantine:2015}. Nevertheless, we are motivated by the
underlying potential of this approach in reducing the computational effort associated with the
active subspace computation. Moreover, this effort could help advance this approach and further motivate
its usage for a wide variety of high-dimensional applications in the future.

First, we implement the grad-free approach to the 19-dimensional problem and compare our findings with
those obtained using the grad-based approach. Later, in
Section~\ref{sec:app}, we compute the active subspace for a 33-dimensional application using the grad-free
approach. Once again, the strategy is implemented in an iterative manner as shown in~\ref{sub:grad}. 
It is observed that the results from the grad-free approach are consistent with the grad-based approach
as discussed further below.

In the grad-free approach, the model output is first computed at an initial set of $n_1$ samples. An independent
set of $M$ samples is also drawn according to the joint probability distribution of the inputs. For each sample
in $M$, $p$ nearest neighbours in the $n_1$ set are identified. A least-squares fit to model evaluations at
the $p$ samples is performed. The slope vector ($\bm{b}_i$) of the regression-fit is stored separately and the process is
repeated for each sample in $M$. The symmetric and positive semi-definite matrix, $\hat{\mat{C}}$ is hence constructed
using the slope vectors as follows:
%
\be
\hat{\mat{C}} = \frac{1}{M}\sum\limits_{i=1}^{M}\bm{b}_i\bm{b}_i^\top = \hat{\bm{W}}\hat{\bm{\Lambda}}\hat{\bm{W}}^\top
\ee
%
Eigenvalue decomposition of $\hat{\mat{C}}$ yields an initial approximation to the
active subspace. The aforementioned sequence of steps is repeated until $\delta w_{1,j}^{(r)}$ as mentioned
earlier in~\ref{sub:grad} is found to be smaller than $\tau$. An outline of the grad-free approach is
provided in Algorithm~\ref{alg:free1}, and the specific sequence of steps pertaining to the construction of
$\hat{\mat{C}}$ is provided in Algorithm~\ref{alg:free2}.

\bigskip
\begin{breakablealgorithm}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
  \caption{An iterative gradient-free approach for discovering the active subspace}
  \begin{algorithmic}[1]
\Require $\theta_l$, $\theta_u$, $M$, $\tau$. 
\Ensure $\Lambda$, $W$, $\alpha$ %$\eta$. 
    \Procedure{Gradient-free}{}
    \State Set $r$ = 0
	\State Draw $n_r$ random samples, $\{\bm{\xi}_k\}_{k=1}^{n_1}$ $\in$ [-1,1]
         according to $\bm{f(\xi)}$. Set $N_\text{total}$ = $n_r$ 
	\State Project to the physical space:
        $\{\bm{\theta}_k\}_{k=1}^{n_r}=\theta_l+0.5(\theta_u-\theta_l)\{\bm{\xi}_k\}_{k=1}^{n_r}$
     \State Draw $M$ random samples, $\{\bm{\nu}_i\}_{i=1}^{M}$ $\in$ [-1,1]
        according to $\bm{f(\xi)}$.
	\State Compute $G(\bm\theta_k)$, $k=1, \ldots, N_\text{total}$.
	\State Choose an integer $p \leq N$ 
	\State For each $i=1, \ldots, M$, compute 
	\[
	\begin{aligned}
	\Phi_i &= \{ p \text{ nearest points in } \{\bm{\xi}_k\}_{k=1}^{N} \text{ to } \bm{\nu}_i\}\\
	\vspace{-2mm}
	\Psi_i &= \text{subset of } G(\bm\theta_k) \text{ corresponding to the points in } \Phi_i
	\end{aligned}
	\]
	\State Compute linear least square fits 
	$G(\bm\theta_k) \approx c_i + b_i^T\bm{\xi}_k$, \  $\bm{\xi}_k \in \Phi_i, \ G(\bm\theta_k) \in \Psi_i$
	\State Compute the matrix 
	\[
	\mat{C} \approx \frac{1}{M} \sum_{i=1}^{M} b_i b_i^T = {W}{\Lambda}{W}^T
	\]
	%\State Eigenvalue decomposition, $\mat{C}$ = $W^{(r)}\Lambda^{(r)} W^{(r)\top}$
	\State Partition the Eigenpairs: $\Lambda^{(r)}~=~ 
        \begin{bmatrix} \Lambda_1^{(r)} & \\ & \Lambda_2^{(r)} \end{bmatrix}$, 
        $W^{(r)}~=~\begin{bmatrix} W_1^{(r)} & W_2^{(r)} \end{bmatrix}$, 
        $\Lambda_1^{(r)}\in \mathbb{R}^{d\times\mathcal{S}}$
	\Loop
		\State Set $r$ = $r$ + 1
		\State Draw $n_r =  \beta n_{r-1}$  new random samples 
		$\{\bm{\xi}_k\}_{k=1}^{n_r}$ $\in$ [-1,1], $\beta =??$ \iffalse $k = n_{r-1}+1,\ldots,n_{r-1}+n_r$. \fi
        \State $N_\text{total}$ = $N_\text{total}$ + $n_r$   
        \State Repeat steps 4-11  
		\iffalse \State Project $\bm{\xi}_k$~$\rightarrow$~$\bm{\theta}_k$.

		\State Compute $G(\bm\theta_k)$, $k=n_{r-1}+1, \ldots, n_{r-1}+n_r$.  
		\State Construct the matrix, $\mat{C}$.
		\State Eigenvalue decomposition, $\mat{C}$ = $W^{(r)}\Lambda^{(r)} W^{(r)\top}$
		\fi
		\State Compute $\delta W_{1,j}^{(r)}$ = 
                       \scalebox{1.25}{$\frac{\|(W_{1,j}^{r})^2 - 
                       (W_{1,j}^{r-1})^2\|}{\|(W_{1,j}^{r-1})^2\|}$}, 
                       $j = 1,\ldots,\mathcal{S}$.
		\If {$\max\left(\delta W_{1,j}^{(r)}\right)<\tau$}
			\State break
		\EndIf
	\EndLoop
	\State Compute $\alpha_i(N_\text{total}) = \sum\limits_{j=1}^{\mathcal{S}} \Lambda_{1_{i,j}}W_{1_{i,j}}^2$,
	$i=1,\ldots,d$.
    \EndProcedure
  \end{algorithmic}
  \label{alg:free1}
\end{breakablealgorithm}
\bigskip

@@

The grad-free approach when implemented to the 19-dimensional H$_2$/O$_2$ reaction kinetics problem
with uncertain pre-exponents ($A_i$'s) also yielded a 1-dimensional active subspace. In Figure~\ref{fig:comp} (left),
we provide a comparison of the components of the dominant eigenvector, obtained using the two approaches.
Additionally, the quantity, $\delta w_{1,j}^{(r)}$ is also plotted in
Figure~\ref{fig:comp} (right) to illustrate convergence characteristics for the grad-free approach.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/eigv6}
\caption{Left: An illustrative comparison of individual components of the converged dominant eigenvector obtained
using the two approaches discussed in~\ref{sub:grad} and~\ref{sub:gradfree}. Right: The quantity,  $\delta w_{1,j}^{(r)}$
is plotted for successive iterations to illustrate the convergence behavior.}
\label{fig:comp}
\end{center}
\end{figure}
%
Individual components of the dominant eigenvector from the two approaches are observed to be in excellent
agreement with each other. Moreover, as expected,  $\delta w_{1,j}^{(r)}$ is observed to decrease with iterations, and
the eigenvector is considered to have converged if its value is found to be smaller than 0.1. Note that only 30 samples
were found to be sufficient for approximating the active subspace in the 19-dimensional input domain.

As mentioned earlier, the model output $f(\bm{\xi}(\bm{\theta}))$ predominantly varies in the active subspace. Hence, 
$f(\bm{\xi}(\bm{\theta}))$ can be transformed as a function of the active variable as $G(\bm{w}_1^\top\bm{\xi})$ in
the subspace. The plot of $G$ versus $\bm{w}_1^\top\bm{\xi}$, regarded as the \textit{sufficient summary plot} (SSP) is 
compared for the two approaches in Figure~\ref{fig:comp_ssp}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.48\textwidth]{./Figures/comp_ssp}
\caption{An illustrative comparison of the SSPs generated using the grad-based and the grad-free approaches.}
\label{fig:comp_ssp}
\end{center}
\end{figure}
%
The two SSPs are observed to be in excellent agreement with each other. Moreover, it is interesting to note
that the variability in ignition
delay based on the considered prior marginals for $A_i$'s is observed to be approximately linear in the
active subspace; consistent with the fact the it is 1-dimensional.

@@

The above comparisons establish consistency between the two approaches. Note that the computational
effort in the case of grad-free approach is significantly smaller. Hence, we are encouraged to use the
grad-free approach for the 33-dimensional application in the next section.

>>>>

In this section, we present the computational strategy used for computing the active subspace in the case of
H$_2$/O$_2$ reaction kinetics with uncertain pre-exponent of individual reaction rates provided in
Table~\ref{tab:kinetics}. Specifically, we employ two strategies: One that involves computation of model
gradients in order to construct the matrix (gradient-based), $\hat{\mat{C}}$ in~\eqref{eq:chat}, and the other that involves a local 
linear approximation of the model output to circumvent the computational cost associated with gradient
computation (gradient-free). The gradient-based approach is based on Algorithm 1.1, and the gradient-free approach is
based on Algorithm 1.2 in~\cite{Constantine:2015}. However, in this work, we have implemented the
two strategies in an iterative manner and have demonstrated potential for additional savings in both cases.
Additionally, for the present application, our findings based on the two strategies are observed to be consistent.
Since, the gradient-free approach is relatively more efficient, this observation encourages its use for high-dimensional
applications as discussed later in Section~\ref{sec:app}. Note however, that the model output is required to be
differentiable throughout the input domain in both cases.

\subsection{Gradient-based approach}
\label{sub:grad}

As discussed earlier, in the gradient-based approach, the elements of $\hat{\mat{C}}$ are estimated using model gradients.
 In situations where the gradient is not available analytically, one could consider using
finite difference and other relatively more efficient techniques such as automatic differentiation~\cite{Kiparissides:2009} 
and adjoint-based gradient computation~\cite{Jameson:1988,Gunzburger:2003,Borzi:2011,Alexanderian:2017}.
Model evaluations at neighbouring points are required if finite difference is used. Hence, for 
$N$ samples in a $d$-dimensional parameter space, $N(d+1)$ model evaluations are needed.

We begin by evaluating the gradient of the model output, $\nabla_{\bm{\xi}}f$, at an initial set of $n_1$ samples. 
For this purpose, $\bm{\xi}_i$'s are projected to the physical space as $\bm{\theta}_i$'s. Using the gradient
evaluations, the matrix, $\hat{\mat{C}}$ is computed. Eigenvalue decomposition of $\hat{\mat{C}}$ yields an initial
estimate of the dominant eigenspace, $\hat{\mat{W}}_1$ and corresponding set of eigenvalues, $\hat{\mat{\Lambda}}_1$.
At each subsequent iteration, model evaluations are generated at a new set of $n_s$ samples. The new set
of gradient evaluations are augmented with the available set to re-construct $\hat{\mat{C}}$ followed by its eigenvalue
decomposition. The relative
change in the norm of the difference in squared value of individual components of the dominant eigenvectors between subsequent 
iterations is evaluated. The process is terminated and the resulting eigenspace is considered to be converged once
highest relative change at iteration $k$, $\delta \hat{\mat{W}}_{1,j}^{(k)}$, is smaller a given tolerance, $\tau$. 
A regression fit to
 $G(\hat{\mat{W}}_1^\top\bm{\xi})$ is used as a surrogate to characterize and quantify the uncertainty in the model
 output. Moreover, the components of the dominant eigenvectors are used to compute the activity scores,
 $\bm{\nu}_r(f)$, which provide an insight into the relative importance of the uncertain inputs. Note that the
 index, $r$, corresponds to the number of dominant eigenvectors in $\hat{\mat{W}}_1$. The sequence of 
 steps for the gradient-based approach are provided in Algorithm~\ref{alg:grad}.

@@

\bigskip
\begin{breakablealgorithm}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
  \caption{An iterative gradient-based approach for discovering the active subspace}
  \begin{algorithmic}[1]
\Require $\theta_l$, $\theta_u$, $\tau$. 
\Ensure $\hat{\mat{\Lambda}}$, $\hat{\mat{W}}$, $\bm{\nu}_r(f)$ %$\eta$. %
    \Procedure{Gradient-based}{}
    \State Set $k$ = 0
	\State Draw $n_k$ random samples, $\{\bm{\xi}_i\}_{i=1}^{n_k}$ 
         according to $\pi$. 
    \State Set $N_\text{total}$ = $n_k$ 
	\iffalse \State Project to the physical space:
        $\{\bm{\theta}_k\}_{k=1}^{n_r}=\theta_l+0.5(\theta_u-\theta_l)\{\bm{\xi}_k\}_{k=1}^{n_r}$ \fi
	\State For each $i=1, \ldots, N_\text{total}$, compute $f(\bm{\xi}_i)$ and the gradient $\bm{g}^i = \nabla_{\bm{\xi}}f(\bm{\xi}_i)$
    \iffalse  
	\Statex\hspace{5mm} Using Finite Difference: 
	\Statex\hspace{5mm} i. Assign a small increment, $d\xi$.
	\Statex\hspace{5mm} ii. Augment the set of samples with neighboring points.
	\be \{\bm{\Xi}_k\}_{k=1}^{n_1(N_p+1)}:~\{\bm{\xi}_k\}_{k=1}^{n_1} \cup
        \{\xi_{k,j}+d\xi\}_{j=1}^{N_p} \nonumber
	\ee
	\Statex\hspace{5mm} iii. Project to the physical space:
        $\{\bm{\theta}_k\}_{k=1}^{n_1(N_p+1)}=\theta_l+0.5(\theta_u-\theta_l)\{\bm{\Xi}_k\}_{k=1}^{n_1(N_p+1)}$
	\Statex\hspace{5mm} iv. Using the augmented set, $\{\bm{\theta}_k\}_{k=1}^{n_1(N_p+1)}$
         compute $\bm{g}^i$.
         \fi 
	\State Compute $\hat{\mat{C}}$ and its eigenvalue decomposition 
		$\hat{\mat{C}}$= $\frac{1}{N_\text{total}}\sum\limits_{i=1}^{N_\text{total}}[\bm{g}^i][\bm{g}^i]^\top$ = 
		$\hat{\mat{W}}^{(k)}\hat{\mat{\Lambda}}^{(k)} \hat{\mat{W}}^{(k)\top}$
	%\State Eigenvalue decomposition, $\mat{C}$ = $W^{(0)}\Lambda^{(0)} W^{(0)\top}$%
	\State Partition: $\hat{\mat{\Lambda}}^{(k)}=
        \begin{bmatrix} \hat{\mat{\Lambda}}_1^{(k)} & \\ & \hat{\mat{\Lambda}}_2^{(k)} \end{bmatrix}$, 
        $\hat{\mat{W}}^{(k)}=\begin{bmatrix} \hat{\mat{W}}_1^{(k)} & \hat{\mat{W}}_2^{(k)} \end{bmatrix}$, 
        $\hat{\mat{\Lambda}}_1^{(k)}\in \mathbb{R}^{N_p\times r}$
	\Loop
		\State Set $k$ = $k$ + 1
		\State Draw $n_k =  \lceil\beta n_{k-1}\rceil$  new random samples 
                $\{\bm{\xi}_i\}_{i=1}^{n_k}$  $\beta\in[0,1]$  $k = n_{k-1}+1,\ldots,n_{k-1}+n_k$.

	%	\State Project $\bm{\xi}_k$~$\rightarrow$~$\bm{\theta}_k$.%
		\State Set $N_\text{total}$ = $N_\text{total}$ + $n_k$ 
		\State Compute $\bm{g}^i = \nabla_{\bm{\xi}_i}f(\bm{\xi}_i)$, 
             	$i=n_{k-1}+1, \ldots, n_{k-1}+n_k$.  
		\State Compute $\hat{\mat{C}}$ = 
        	$\frac{1}{N_\text{total}}\sum\limits_{k=1}^{N_\text{total}}[\bm{g}^i][\bm{g}^i]^\top$
		\State Eigenvalue decomposition, $\hat{\mat{C}}$ = $\hat{\mat{W}}^{(k)}\hat{\mat{\Lambda}}^{(k)}
		 \hat{\mat{W}}^{(k)\top}$
		\State Compute $\delta \hat{\mat{W}}_{1,j}^{(k)}$ = 
                       \scalebox{1.25}{$\frac{\|(\hat{\mat{W}}_{1,j}^{k})^2 - 
                       (\hat{\mat{W}}_{1,j}^{k-1})^2\|}{\|(\hat{\mat{W}}_{1,j}^{k-1})^2\|}$}, 
                       $j = 1,\ldots,r$.
		\If {$\max\left(\delta \hat{\mat{W}}_{1,j}^{(k)}\right)<\tau$}
			\State break
		\EndIf
	\EndLoop
	\State Compute $\nu_{i,r}(f) = \sum\limits_{j=1}^{r} \hat{\mat{\Lambda}}_{i,j}\hat{\mat{W}}_{i,j}^2$,
	$i=1,\ldots,N_p$.

@@

\subsection{Gradient-free approach}
\label{sub:gradfree}

As mentioned earlier, the gradient-free approach does not involve computation of the gradient of the model
output. Instead, a least-squares fit is performed to the local linear regression-based approximation
to the model output. While the gradient-free approach offers a significant potential for additional 
computational savings, its ability to reasonably approximate the active subspace 
has not been investigated for a wide variety of problems~\cite{Constantine:2015}. Nevertheless, we are motivated by the
underlying potential of this approach in reducing the computational effort associated with the
active subspace computation. Moreover, this effort could help advance this approach and further motivate
its usage for a wide variety of high-dimensional applications in the future.

First, we implement the gradient-free approach to the 19-dimensional problem and compare our findings with
those obtained using the gradient-based approach. Later, in
Section~\ref{sec:app}, we compute the active subspace for a 33-dimensional application using the gradient-free
approach. Once again, the strategy is implemented in an iterative manner as shown in~\ref{sub:grad}. 
It is observed that the results from the gradient-free approach are consistent with the gradient-based approach
as discussed further below.

In the gradient-free approach, the model output is first computed at an initial set of $n_1$ samples. An independent
set of $M$ samples is also drawn according to the joint probability distribution of the inputs. For each sample
in $M$, $p$ nearest neighbours in the $n_1$ set are identified. A least-squares fit to model evaluations at
the $p$ samples is performed. The slope vector ($\bm{b}_i$) of the regression-fit is stored separately and the process is
repeated for each sample in $M$. The symmetric and positive semi-definite matrix, $\hat{\mat{C}}$ is hence constructed
using the slope vectors as follows:
%
\be
\hat{\mat{C}} = \frac{1}{M}\sum\limits_{i=1}^{M}\bm{b}_i\bm{b}_i^\top = \hat{\bm{W}}\hat{\bm{\Lambda}}\hat{\bm{W}}^\top
\ee
%
Eigenvalue decomposition of $\hat{\mat{C}}$ yields an initial approximation to the
active subspace. The aforementioned sequence of steps is repeated until $\delta w_{1,j}^{(r)}$ as mentioned
earlier in~\ref{sub:grad} is found to be smaller than $\tau$. An outline of the gradient-free approach is
provided in Algorithm~\ref{alg:free1}, and the specific sequence of steps pertaining to the construction of
$\hat{\mat{C}}$ is provided in Algorithm~\ref{alg:free2}.

\bigskip
\begin{breakablealgorithm}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
  \caption{An iterative gradient-free approach for discovering the active subspace}
  \begin{algorithmic}[1]
\Require $\theta_l$, $\theta_u$, $M$, $\tau$. 
\Ensure $\Lambda$, $W$, $\alpha$ %$\eta$. 
    \Procedure{Gradient-free}{}
    \State Set $k$ = 0
	\State Draw $n_k$ random samples, $\{\bm{\xi}_j\}_{j=1}^{n_k}$
	\State Set $N_\text{total}$ = $n_k$ 
	\iffalse \State Project to the physical space:
        $\{\bm{\theta}_k\}_{i=1}^{n_k}=\theta_l+0.5(\theta_u-\theta_l)\{\bm{\xi}_k\}_{k=1}^{n_r}$
        \fi
     \State Draw $M$ random samples, $\{\bm{\nu}_i\}_{i=1}^{M}$ 
	\State Compute $f(\bm\xi_j)$, $j=1, \ldots, N_\text{total}$.
	\State Compute the matrix $\mathbb{C}$ and its eigenvalue decomposition using Algorithm 3
	%\State Eigenvalue decomposition, $\mat{C}$ = $W^{(r)}\Lambda^{(r)} W^{(r)\top}$
	\State Partition: $\Lambda^{(k)}~=~ 
        \begin{bmatrix} \Lambda_1^{(k)} & \\ & \Lambda_2^{(k)} \end{bmatrix}$, 
        $W^{(k)}~=~\begin{bmatrix} W_1^{(k)} & W_2^{(k)} \end{bmatrix}$, 
        $\Lambda_1^{(k)}\in \mathbb{R}^{d\times\mathcal{S}}$
	\Loop
		\State Set $k$ = $k$ + 1
		\State Draw $n_k =  \beta n_{k-1}$  new random samples 
		$\{\bm{\xi}_j\}_{j=1}^{n_k}$  $\beta =??$  $j = n_{k-1}+1,\ldots,n_{k-1}+n_k$. 
        \State $N_\text{total}$ = $N_\text{total}$ + $n_k$    
	\iffalse	\State Project $\bm{\xi}_k$~$\rightarrow$~$\bm{\theta}_k$. \fi

		\State Compute $f(\bm\xi_j)$, $j=n_{k-1}+1, \ldots, n_{k-1}+n_k$.  
		\State Construct the matrix, $\mat{C}$ and its eigenvalue decomposition
		\State Compute $\delta W_{1,j}^{(k)}$ = 
                       \scalebox{1.25}{$\frac{\|(W_{1,j}^{k})^2 - 
                       (W_{1,j}^{k-1})^2\|}{\|(W_{1,j}^{k-1})^2\|}$}, 
                       $j = 1,\ldots,\mathcal{S}$.
		\If {$\max\left(\delta W_{1,j}^{(k)}\right)<\tau$}
			\State break
		\EndIf
	\EndLoop
	\State Compute $\alpha_i(N_\text{total}) = \sum\limits_{j=1}^{\mathcal{S}} \Lambda_{1_{i,j}}W_{1_{i,j}}^2$,
	$i=1,\ldots,d$.
    \EndProcedure
  \end{algorithmic}
  \label{alg:free1}
\end{breakablealgorithm}
\bigskip

@@

The gradient-free approach when implemented to the 19-dimensional H$_2$/O$_2$ reaction kinetics problem
with uncertain pre-exponents ($A_i$'s) also yielded a 1-dimensional active subspace. In Figure~\ref{fig:comp} (left),
we provide a comparison of the components of the dominant eigenvector, obtained using the two approaches.
Additionally, the quantity, $\delta w_{1,j}^{(r)}$ is also plotted in
Figure~\ref{fig:comp} (right) to illustrate convergence characteristics for the gradient-free approach.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/eigv6}
\caption{Left: An illustrative comparison of individual components of the converged dominant eigenvector obtained
using the two approaches discussed in~\ref{sub:grad} and~\ref{sub:gradfree}. Right: The quantity,  $\delta w_{1,j}^{(r)}$
is plotted for successive iterations to illustrate the convergence behavior.}
\label{fig:comp}
\end{center}
\end{figure}
%
Individual components of the dominant eigenvector from the two approaches are observed to be in excellent
agreement with each other. Moreover, as expected,  $\delta w_{1,j}^{(r)}$ is observed to decrease with iterations, and
the eigenvector is considered to have converged if its value is found to be smaller than 0.1. Note that only 30 samples
were found to be sufficient for approximating the active subspace in the 19-dimensional input domain.

As mentioned earlier, the model output $f(\bm{\xi}(\bm{\theta}))$ predominantly varies in the active subspace. Hence, 
$f(\bm{\xi}(\bm{\theta}))$ can be transformed as a function of the active variable as $G(\bm{w}_1^\top\bm{\xi})$ in
the subspace. The plot of $G$ versus $\bm{w}_1^\top\bm{\xi}$, regarded as the \textit{sufficient summary plot} (SSP) is 
compared for the two approaches in Figure~\ref{fig:comp_ssp}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.48\textwidth]{./Figures/comp_ssp}
\caption{An illustrative comparison of the SSPs generated using the gradient-based and the gradient-free approaches.}
\label{fig:comp_ssp}
\end{center}
\end{figure}
%
The two SSPs are observed to be in excellent agreement with each other. Moreover, it is interesting to note
that the variability in ignition
delay based on the considered prior marginals for $A_i$'s is observed to be approximately linear in the
active subspace; consistent with the fact the it is 1-dimensional.

@@

The above comparisons establish consistency between the two approaches. Note that the computational
effort in the case of gradient-free approach is significantly smaller. Hence, we are encouraged to use the
gradient-free approach for the 33-dimensional application in the next section.
